% see this link for more info:
% https://github.com/tudace/tuda_latex_templates/blob/master/example/DEMO-TUDaThesis.tex
\documentclass[
	english,
	class=report,custommargins=true,marginpar=false,
	accentcolor=9c,% red-y emphasis color
	thesis={type=bachelor},% bachelor vs master
	%BCOR=5mm,%binding correction
	fontsize=11pt,% font size
%	logofile=example-image, %Falls die Logo Dateien nicht vorliegen
]{tudapub}

%%My own stuff
\usepackage{listings}
	\definecolor{codegray}{rgb}{0.5,0.5,0.5}
	\definecolor{backcolour}{rgb}{0.98,0.98,0.96}
	\definecolor{stringcolour}{RGB}{98,151,85}
	\definecolor{keywordcolour}{RGB}{204,120,50}
	% \definecolor{keywordcolour}{RGB}{0,0,150}
	\lstdefinestyle{jetbrains}{
		% backgroundcolor=\color{backcolour},   
		commentstyle=\color{stringcolour},
		keywordstyle=\color{keywordcolour},
		numberstyle=\tiny\color{codegray},
		stringstyle=\color{stringcolour},
		basicstyle=\ttfamily\footnotesize,
		breakatwhitespace=false,         
		breaklines=true,                 
		captionpos=b,                    
		keepspaces=true,                 
		numbers=left,                    
		numbersep=5pt,                  
		showspaces=false,                
		showstringspaces=false,
		showtabs=false,                  
		tabsize=2
	}
	\lstset{
		style=jetbrains,
		language=scala,
		morekeywords={end},
		escapebegin=\color{stringcolour}
	}
\usepackage{scrhack} % to prevent an error when using a self defined listing command
% math expressions in 
\usepackage{xcolor}% http://ctan.org/pkg/listings
\usepackage[framemethod=TikZ]{mdframed}% http://ctan.org/pkg/mdframed
\usepackage{etoolbox}{% http://ctan.org/pkg/etoolbox
	\mdfdefinestyle{listingstyle}{%
		backgroundcolor=backcolour,
		outerlinewidth=0.25pt,
		outerlinecolor=black,
		innerleftmargin=13pt,
		innerrightmargin=5pt,
		innertopmargin=0pt,
		innerbottommargin=0pt
	}
	\BeforeBeginEnvironment{lstlisting}{\begin{mdframed}[style=listingstyle]}
	\AfterEndEnvironment{lstlisting}{\end{mdframed}}

\usepackage{tikz-cd}
	\usetikzlibrary{babel} % to fix error with "" in arrow labels

\newcommand{\lstscala}[3]{\lstinputlisting[firstline=#2, lastline=#3]{listings/#1}}
\newcommand{\lstscalacaption}[5]{\lstinputlisting[firstline=#2, lastline=#3, label={#5}, caption={#4}]{listings/#1}}
\newcommand{\code}[1]{\texttt{#1}}

\newcommand{\pluseq}{\mathrel{+}\mathrel{\mkern-2mu}=}


%%This is needed for pdfTeX version older than April 2018
\usepackage{iftex}
\ifPDFTeX
\usepackage[utf8]{inputenc}
\fi

\usepackage[main=english, ngerman]{babel} % language support
\usepackage[autostyle]{csquotes}          % quote-characters e.g. " "
\usepackage{microtype}                    % microtyping
\usepackage{biblatex}                     % bibliography

\bibliography{content/references}
%\addbibresource{content/references.bib}


%%latex package recommendations: tables + math
\usepackage{tabularx}   % tables columns x to expand and fill linewidth
%\usepackage{longtable} % multi-page tables
%\usepackage{xltabular} % both features from tabularx and longtable
\usepackage{booktabs}   % better looking style for table rulers
\usepackage{mathtools} % better than amsmath
\usepackage{amssymb}   % more symbols
\usepackage{siunitx}   % si-units

\usepackage{esdiff}

\usepackage{caption}
	% \captionsetup[lstlisting]{labelfont={bf}, labelsep=colon, justification=centering}

\newcommand{\reflst}[1]{\hyperref[#1]{listing} \ref{#1}} % reference listings
\newcommand{\reffig}[1]{\hyperref[#1]{figure} \ref{#1}} % reference figures
\newcommand{\refsec}[1]{\hyperref[#1]{section} \ref{#1} (\nameref{#1})} % reference sections



\newcommand{\todocolor}[2]{{\color{#1}{\textbf{TODO: }#2}}}
\newcommand{\todo}[1]{\todocolor{red}{#1}}
\newcommand{\todogrammar}{\todocolor{orange}{Englisch spaek yu kan¿}}
\newcommand{\todopunctuation}{\todocolor{teal}{,.?!}}
\newcommand{\todowording}{\todocolor{green}{wording?}}
\newcommand{\todocite}[1]{\todocolor{purple}{Cite #1}}

\newcommand{\R}{\mathbb{R}}




\begin{document}

\Metadata{
	title=Comparing implementation strategies for differentiable programming,
	author=Daniel Stricker
}

\title{Comparing implementation strategies for differentiable programming}
\subtitle{}
\author{Daniel Stricker}%optional argument used in affidavit
\reviewer{Prof. Dr.-Ing. Mira Mezini \and David Richter, M.Sc.}
\department{inf}
%\institute{Software Technology}
\group{Software Technology}

\submissiondate{\today}
\examdate{\today}

%\tuprints{urn=1234,printid=12345}
%\dedication{...}

\maketitle

\affidavit

% \input{content/abstract}
% \chapter{Abstract [1]}

\tableofcontents

\input{content/introduction}
\input{content/implementation/implementation}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Planned table of contents
% \IMRADlabel{introduction}
% \input{content/introduction}
\chapter{Introduction [1/2]}
\section{Motivation, Why? -> ML needs this for Newton etc}
% \IMRADlabel{methods}
\section{Overview}
\subsection{Short summary of all implementation styles (i.e. CPS, Monad, tape, macro) [1/2]}
\section{Details}
\subsection{Example backpropagation by hand [1]}
\subsection{Go through a real ML task (the implemented test) with "grad" as blackbox -> defines "goal" [1]}
% \input{content/implementation}

\chapter{Implementation}
\section{Forward mode [1]}
\subsection{by operator overloading (including macros)}
\subsection{Match type implementation (and bad performance). Explain type level calculations (Peano). Pro/Con [1]}

\section{Backpropagation with mutation}
\subsection{Tape [1/2]} %maybe first?
\subsection{CPS}
\subsection{Explain Continuations (without shift/reset) [1/2]}
\subsection{Encode backward pass as continuation [1/2]}
\subsection{MonadCPS}
\subsection{ -- For-comprehensions (flatMap, map) [1/2]}
\subsection{ -- First try: .map() "closes" monad [1/2] }
\subsection{ -- Second try: Chaining of functions with global tape [1/2]}

\section{Backpropagation without mutation (functional)}
\subsection{Why functional? (state, multiple calculations, reasoning) [1/2]}
\subsection{CPS, MonadCPS [1]}
\subsection{Tape}
\subsection{ -- First try: naive (val, var difference -> multiple updates) [1/2]}
\subsection{ -- Second try: Every node has to be calculated once [1]}
\subsection{Con: (speed, ...) [1]}

\section{Macros? [2]}


% \IMRADlabel{results}
\chapter{Evaluation}
\section{List all comparison metrics (Runtime, LOC, compile time) [1]}
\section{Interpretation [1/2]}
\chapter{Related work}
\section{lantern paper [1/2]}

% \IMRADlabel{discussion}
\chapter{Conclusion [1/2]}



% Pages: 15
\printbibliography

\end{document}
